Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-02-02 causal-bandit/icml-2015/general.tex
started to clean up introduction
started to clean up Sec 3

\newcommand{\calP}{\mathcal P}
\newcommand{\calA}{\mathcal A}
\newcommand{\x}{\boldsymbol{x}}
\newcommand{\Ps}{\operatorname{P}}

@@

\begin{algorithm}[H]
\caption{General Algorithm}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $T$, $\eta \in [0,1]^{|\calA|}$, $B \in [0,\infty)^{|\calA|}$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $I_t = a$ with probability $\eta$
\STATE Do action $I_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\STATE For each $a \in \calA$ compute an estimate of its return:
\eq{
\forall a\in \calA \qquad \hat \mu_a = \frac{1}{T} \sum_{t=1}^T Z_a(X_t)
}
\STATE Choose $I = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}

@@

\begin{proof}
By definition, $m(\eta^*) \leq m(\eta)$ for all $\eta$. Let $\eta_a = 1/|\calA|$. Then
\eq{
m(\eta) 
&= \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \\
&\leq \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] \\
&\leq \max_a \EE_{\Ps_a}\left[\frac{1}{\eta_a}\right] = |\calA|
}
as required.
\end{proof}

The second observation is that in the parallel bandit setting the $m$ given in this section coincides with the $m$ given there.
Note that in the previous section we did not assume knowledge of the conditional distributions on $X$ and instead only the graph structure.
Thus the algorithm given in that setting has broader applicability to that particular graph than the generic algorithm given here.
\todot{Finish this}

Finally, we believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to minimax optimal, but leave lower bounds
for future work.
\todot{Mention obvious modification to deal with large gaps}

>>>>

\newcommand{\calP}{\mathcal P}
\newcommand{\x}{\boldsymbol{x}}
\newcommand{\Ps}{\operatorname{P}}

@@

\begin{algorithm}[H]
\caption{General Algorithm}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $T$, $\eta \in [0,1]^{|\calA|}$, $B \in [0,\infty)^{|\calA|}$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $a_t = a$ with probability $\eta$
\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\STATE For each $a \in \calA$ compute an estimate of its return:
\eq{
\forall a\in \calA \qquad \hat \mu_a = \frac{1}{T} \sum_{t=1}^T Z_a(X_t)
}
\STATE {\bf return} $\hat a^* = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}

@@

\begin{proof}
By definition, $m(\eta^*) \leq m(\eta)$ for all $\eta$. Let $\eta_a = 1/|\calA|$. Then
\eq{
m(\eta) 
&= \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \\
&\leq \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] \\
&\leq \max_a \EE_{\Ps_a}\left[\frac{1}{\eta_a}\right] = |\calA| \qedhere
}
\end{proof}

\todo{give reference}
The second observation is that in the parallel bandit setting the $m(\eta)$ given in this section coincides with the $m$ in Equation blah.
Recall in that setting that 
\eq{
\actions = \set{do()} \cup \set{do(X_i = j) : 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
}
For $a = do()$ we let $\eta_a = 1/2$. Assume without loss of generality that $\P{X_i = 1} \geq \P{X_i = 0}$ for all $i$ and
$\P{X_i = 0}$ is monotone increasing in $i$. Then let $m = \max\set{i : \P{X_i = 0} \leq 1 / i}$, which is the same as in \S3. 
For $i \leq m$ and $a = do(X_i = 0)$ let $\eta_a = 1/(2m)$. Then it is easy to check that $m(\eta) \leq 2m$.

\begin{remark}
The algorithm in the previous section for the parallel bandit problem did not assume knowledge of the conditional distributions on $X$ and 
instead only the graph structure. Thus the algorithm given in that setting has broader applicability to that particular graph than the generic algorithm given here.
Finally, we believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to minimax optimal, but leave lower bounds
for future work.
\end{remark}

### 2016-02-02 causal-bandit/icml-2015/setup.tex
started to clean up Sec 3

A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$ and a set of \emph{allowed interventions} $\mathcal{A}$.
One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
The following game is then played over $T$ rounds.
In round $t$, the learner \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ then \emph{observes} sampled values for all non-intervened variables $\vec{X}^c_t$ drawn from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$, including the \emph{reward} $Y_t \in \{0,1\}$.

Although we will focus on the intervene-then-observe ordering of events within each round, other scenarios are also possible. 
If the non-intervened variables are observed before an intervention is selected our framework reduces to stochastic contextual bandits, which are already well understood~\citep{Agarwal2014}. 
Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of 
allowable interventions thereby reducing the complexity.

The learning goal we will consider and analyse is identifying the intervention with highest expected reward by round $T$.
This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009} or ``best-arm identification'' problem~\citep{Gabillon2012}.
This objective is most appropriate when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. 
We denote the expected reward for the intervention $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.
Performance for the first goal can be measured by \emph{simple regret}, \ie, the gap between the optimal expected reward and the intervention $a_T$ chosen by the learner on round $T$:
\begin{equation}
\label{eq:regret-simple}
	\simpleregret = \mu^* - \E{\mu_{a_T}}.
\end{equation}

Another common objective in bandit problems is maximising the \emph{cumulative regret}, the total expected reward over all $T$ rounds. 
We have established upper bounds for cumulative regret for parallel bandits but not the general causal bandit case.
Due to space limitations we focus on simple regret and defer these to Appendix~\ref{app:cumulatve}.

@@

In the remainder of this paper we focus on a causal model that is sufficient to demonstrate the separation of regret bounds between algorithms that make use of the additional causal information and those which do not.

\subsection{Exploiting the Causal Model}

@@

>>>>

A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$ and a set of \emph{allowed interventions} $\mathcal{A}$.
One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
We denote the expected reward for the intervention $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and 
the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.

The game proceeds over $T$ rounds.
In round $t$, the learner \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ based on previous observations. 
It then \emph{observes} sampled values for all non-intervened variables $\vec{X}^c_t$ drawn from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$, 
including the \emph{reward} $Y_t \in \{0,1\}$. 
After $T$ observations the learner outputs an estimate of the optimal action $\hat a^*_T$.

The objective of the learner is to minimise the simple regret
\begin{equation}
\label{eq:regret-simple}
	\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.
\end{equation}
This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009} or ``best-arm identification'' problem~\citep{Gabillon2012} and
is most appropriate when the learner has a fixed learning budget after which its policy will be fixed indefintely.

Although we will focus on the intervene-then-observe ordering of events within each round, other scenarios are possible. 
If the non-intervened variables are observed before an intervention is selected our framework 
reduces to stochastic contextual bandits, which are already reasonably well understood~\citep{Agarwal2014}. 
Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of 
allowable interventions thereby reducing the complexity.

@@

\iffalse
\subsection{Exploiting the Causal Model}

@@

\fi

### 2016-02-02 causal-bandit/icml-2015/simple-regret.tex
started to clean up Sec 3

In this section we propose and analyse an algorithm for achieving the optimal simple regret in a special case of a causal bandit problem.

\subsection{The Parallel Bandit Problem}

We now consider a class of a causal bandit problems which we call \emph{parallel bandits}.
It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro}, including the farming example. It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.

The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \R$, as shown in Figure~\ref{fig:causalStructure}.
All variables are observable and the set of allowable actions are all size 0 and size 1 interventions: 
\[
	\mathcal{A} = \{do()\} \cup \{ do(X_i = j) \colon i \in \{1, \ldots, N\}, j \in \{0,1\}\}.
\]
In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high). 
In this case, the interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ might represent the use of shades or heat lamps to keep the temperature low or high, respectively.

@@

In each round of the game, the learner either purely observes by selecting $do()$ or chooses a single variable to intervene on and the value to set it to. 
The remainder of the variables are simultaneously set by independently biased coin flips. 
The value of all variables are then used to determine the distribution of rewards for that round.
Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
\todom{Should $Y$ be 0-1 valued or real everywhere?}
The value of the reward variable is distributed as $Y = r(\vec{X}) + \eta$ where $r : \{0,1\}^N \to \R$ is an arbitrary, fixed, and unknown function and $\eta$ is $1$-subgaussian noise with a distribution that may depend on $\vec{X}$.
In the farming example, this choice of $Y$ models the fixed, noisy but unknown dependence of crop yield on the temperature, soil, and moisture.

\todom{$i,j$ notation to $a \in \actions$}
In our analysis below, we will often use the pair $(i,j)$ for $i \in \{1, \ldots, N\}$ and $j \in \{0,1\}$ as a shorthand for the intervention $do(X_i = j)$.
We will also make use of the expansion of the expected reward for $do(X_i = j)$:
\eq{
\mu_{i,j} 
&= \E{r(X)|do(X_i = j)} \\
&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k}\,.  
}
The sum and product use the restrictions $x_i = j$ and $k \ne i$ because the intervention forces $X_i = j$ with probability 1.
The optimal intervention is denoted $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$. 
Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.

@@

\begin{algorithm}[h]
\caption{Causal Best Arm Identification}\label{alg:simple}
\begin{algorithmic}[1]
\STATE {\bf Input:} Total rounds $T$, causal graph $\mathcal{G}$, allowable interventions $\actions$. 
\FOR{$t \in 1,\ldots,(T - 1) / 2$}
\STATE Perform empty intervention $do()$; observe $\vec{x}_t$ and $r_t$
\ENDFOR
\FOR{$a = do(X_i = x) \in \actions$}
\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{x_{t,i} = x}$
\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{x_{t,i} = x} r_t$
\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T-1}$ and $\hat{V}_a = \frac{1}{\hat{p}_a}$
\ENDFOR
\STATE Compute $\hat{m} = m(\hat{\vec{V}})$ and $A = \set{a \in \actions \colon \hat{V}_a \geq \hat{m}}$.
\STATE Let $T_A := \frac{T-1}{2 |A|}$ be times to sample each $a\in A$.
\FOR{$a = do(X_i = x) \in A$}
\FOR{$t \in 1,\ldots,T_A$}
\STATE Intervene with $a$ and observe $r_t$
\ENDFOR
\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} r_t$
\ENDFOR
\RETURN estimated optimal $\hat{a}^* \in \argmax_{a\in\actions} \hat{\mu}_a$
\end{algorithmic}
\end{algorithm}

>>>>

In this section we propose and analyse an algorithm for achieving the optimal regret in a natural special 
case of the causal bandit problem which we call the {\it parallel bandit}.
It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro}, including the farming example. 
It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.

The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a 
reward variable $Y \in \set{0,1}$, as shown in Figure~\ref{fig:causalStructure}.
All variables are observable and the set of allowable actions are all size 0 and size 1 interventions: 
\eq{
	\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
}
In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high). 
In this case, the interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ might represent the use of shades or heat lamps to keep the temperature low or high, respectively.

@@

In each round of the game, the learner either purely observes by selecting $do()$ or chooses a single variable to intervene on and the value to set it to. 
The remainder of the variables are simultaneously set by independently biased coin flips. 
The value of all variables are then used to determine the distribution of rewards for that round.
Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
%\todom{Should $Y$ be 0-1 valued or real everywhere?} YES!
The value of the reward variable is distributed as $\P{Y = 1|\vec{X}} = r(\vec{X})$ where 
$r : \{0,1\}^N \to \R$ is an arbitrary, fixed, and unknown function. 
In the farming example, this choice of $Y$ models the success or failure of a seasons crop, which depends stochastically on the various environment variables.
Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.

%\todom{$i,j$ notation to $a \in \actions$}
%In our analysis below, we will often use the pair $(i,j)$ for $i \in \{1, \ldots, N\}$ and $j \in \{0,1\}$ as a shorthand for the intervention $do(X_i = j)$.
%We will also make use of the expansion of the expected reward for $do(X_i = j)$:
%\eq{
%\mu_{i,j} 
%&= \E{r(X)|do(X_i = j)} \\
%&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k}\,.  
%}
%The sum and product use the restrictions $x_i = j$ and $k \ne i$ because the intervention forces $X_i = j$ with probability 1.
%The optimal intervention is denoted $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$.

@@

The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data.
The fact that $X_1,\ldots,X_N$ are independent can be used to create good estimators of the returns of those
actions $do(X_i = j)$ for which $\P{X_i = j}$ is large. These are the interventions that are essentially observed
from observational data. Those actions where $\P{X_i = j}$ is small may not be observed (often) in the first $T/2$ rounds, so
the estimates of the returns of these actions from the first $T/2$ samples could be quite poor. For the remaining $T/2$ rounds
the algorithm makes interventions on the actions $a = do(X_i = j)$ for which $\P{X_i = j}$ is small. The precise allocation is
carefully chosen to minimise the worst-case simple regret.

\begin{algorithm}[h]
\caption{Parallel Bandit Algorithm}\label{alg:simple}
\begin{algorithmic}[1]
\STATE {\bf Input:} Total rounds $T$ and $N$.
\FOR{$t \in 1,\ldots,T / 2$}
\STATE Perform empty intervention $do()$; observe $\vec{X}_t$ and $Y_t$
\ENDFOR
\FOR{$a = do(X_i = x) \in \actions$}
\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{X_{t,i} = x}$
\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{X_{t,i} = x} Y_t$ \\[0.2cm]
\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$ and $\hat{V}_a = \frac{1}{\hat{p}_a}$
\ENDFOR
\STATE Compute $\hat{m} = m(\hat{\vec{V}})$ and $A = \set{a \in \actions \colon \hat{V}_a \geq \hat{m}}$.
\STATE Let $T_A := \frac{T-1}{2 |A|}$ be times to sample each $a\in A$.
\FOR{$a = do(X_i = x) \in A$}
\FOR{$t \in 1,\ldots,T_A$}
\STATE Intervene with $a$ and observe $r_t$
\ENDFOR
\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} r_t$
\ENDFOR
\RETURN estimated optimal $\hat{a}^* \in \argmax_{a\in\actions} \hat{\mu}_a$
\end{algorithmic}
\end{algorithm}
